\subsection{An Attractive Non-Solution}
\label{sec:cotc}

%The memoization device outlined in \textsection\ref{sec:memdev} can, in principle, also be 
%applied to tagged classes. The dynamic cast will be replaced by a small 
%compile-time template meta-program that checks whether the class associated with 
%the given tag is derived from the target type of the case clause. If so, a static 
%cast can be used to obtain the offset.

%Despite its straightforwardness, we felt that it should be possible to do better 
%than the general solution, given that each class is already identified with a 
%dedicated constant known at compile time.

While Wirth' linked list encoding was considered slow for subtype testing, it can 
be adopted for quite efficient type switching on a class hierarchy with no 
repeated inheritance. The idea is to combine fast switching on closed 
algebraic datatypes with a loop that tries the tags of base classes when 
switching on derived tags fails.

%The nominal subtyping of \Cpp{} effectively gives every class multiple types. The 
%idea is thus to associate with the type not only its most-derived tag, but also 
%the list of tags of all its base classes. In a compiler implementation such a 
%list can be stored inside the virtual table of a class, while in our library 
%solution it is shared between all the instances with the same most-derived tag 
%in a less efficient global map, associating the tag to its tag list.

For simplicity of presentation we assume a pointer to an array of tags be available 
directly through the subject's \code{taglist} data member. The array is of 
variable size: its first element is always the tag of the subject's dynamic 
type, while its end is marked with a dedicated \code{end_of_list} marker, 
distinct from all the tags. The tags in between are topologically sorted 
according to the subtyping relation with incomparable siblings listed in 
\emph{local precedence order} -- the order of the direct base classes used in 
the class definition. The list resembles the \emph{class precedence list} of 
object-oriented descendants of Lisp (e.g. Dylan, Flavors, LOOPS, and CLOS) used 
there for \emph{linearization} of class hierarchies. 
We also assume the tag-constant associated with a class \code{Di} is accessible 
through a static member \code{Di::class_tag}. These simplifications are not 
essential and the library does not rely on any of them.
%Instead, the user can retroactively narrate to the library the specific tag 
%encoding used through a trait-like class.

A type switch, below, %, built on top of a hierarchy of tagged classes, 
proceeds as 
a regular switch on the subject's tag. If the jump succeeds, we found an exact 
match; otherwise, we get into a default clause that obtains the next tag in the
list and jumps back %to the beginning of the switch statement 
for a rematch:

\begin{lstlisting}[keepspaces]
    size_t attempt = 0; 
    size_t tag = subject->taglist[attempt];
ReMatch:
    switch (tag) {
    default:
        tag = subject->taglist[++attempt];
        goto ReMatch;
    case end_of_list: 
        break;
    case D1::class_tag: 
        D1& match = static_cast<D1&>(*subject); s1;
        break;
        ...
    case Dn::class_tag: 
        Dn& match = static_cast<Dn&>(*subject); sn;
        break;
    }
\end{lstlisting}

\noindent
The above structure, which we call a \emph{tag switch}, implements a variation of 
best-fit semantics based on local precedence order. It lets us dispatch to the case 
clause of the most-specialized class with an overhead of initializing two 
local variables, compared to an efficient switch used on algebraic data types. 
Dispatching to a case clause of a base class will take time roughly proportional 
to the distance between the matched base class and the derived class in the 
inheritance graph, thus the technique is not constant. When none of the base 
class tags was matched, we will necessarily reach the end\_of\_list marker %in the list 
and exit the loop. %As mentioned before, 
The default clause, %of the type switch 
again, can be implemented with a case clause on the subject type's tag: \code{case S::class_tag:}

The efficiency of the above code crucially depends on the set of tags 
being small and sequential to justify the use of a jump table instead of a
decision tree to implement the switch. This is usually not a problem in closed 
hierarchies based on tag encoding since the designer of the hierarchy handpicks 
the tags herself. The use of a static cast %to obtain proper reference once the most specialized derived class has been established, 
however, essentially limits the use of 
this mechanism to non-repeated inheritance only. This only refers to the way target 
classes inherit from the subject type -- they can freely inherit from other classes. 
%as long as they inherit the subject type through non-repeated inheritance only. 
Due to these restrictions, the technique is not open because it may  
violate independent extensibility. We discuss in \textsection\ref{sec:cmp} that 
making the technique more open will also eradicate its performance advantages.
