\section{Related Work} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{sec:rw}

\emph{Extensible Visitors with Default Cases}~\cite[\textsection 
4.2]{Zenger:2001} attempt to solve the extensibility problem of visitors; 
however, the solution, after 
remapping it onto C++, has problems of its own. The visitation interface 
hierarchy can easily be grown linearly (adding new cases for the new classes in 
the original hierarchy each time), but independent extensions by different  
authorities require developer's intervention to unify them all, before they can 
be used together. This may not be feasible in environments that use dynamic 
linking. To avoid writing even more boilerplate code in new visitors, the 
solution would require usage of virtual inheritance, which typically has 
an overhead of extra memory dereferencing. On top of the double dispatch already 
present in the visitor pattern, the solution will incur two additional virtual 
calls and a dynamic cast for each level of visitor extension. Additional double 
dispatch is incurred by forwarding of default handling from a base visitor to a 
derived one, while the dynamic cast is required for safety and can be replaced 
with a static cast when the visitation interface is guaranteed to be grown linearly 
(extended by one authority only). Yet another virtual call is required to be 
able to forward computations to subcomponents on tree-like structures to the 
most derived visitor. This last function lets one avoid the necessity of using 
the heap to allocate a temporary visitor through the \emph{Factory Design 
Pattern}~\cite{DesignPatterns1993} used in the \emph{Extensible Visitor} solution 
originally proposed by Krishnamurti, Felleisen and Friedman~\cite{Krishnamurthi98}.

In order to address the expression problem in Haskell, L\"{o}h and Hinze proposed to 
extend its type system with open data types and open functions~\cite{LohHinze2006}.
Their solution allows the user to mark top-level data types and functions as 
open and then provide concrete variants and overloads anywhere in the program. 
Open data types are extensible but not hierarchical, which largely avoids the 
problems discussed here. The semantics of open extension is given by 
transformation into a single module, where all the definitions are seen in one 
place. This is a significant limitation of their approach that prevents it from 
being truly open, since it essentially assumes a whole-program view, which 
excludes any extension via DLLs. As is the case with many other implementations 
of open extensions, the authors rely on the closed world for efficient 
implementation: in their implementation, \emph{``data types can only be entirely 
abstract (not allowing pattern matching) or concrete with all constructors with 
the reason being that pattern matching can be compiled more efficiently if the 
layout of the data type is known completely''}. The authors also believe that 
\emph{there are no theoretical difficulties in lifting this restriction, but it 
might imply a small performance loss if closed functions pattern match on open 
data types}. Our work addresses exactly this problem, showing that it is not 
only theoretically possible but also practically efficient and in application to 
a broader problem.

Polymorphic variants in OCaml~\cite{garrigue-98} allow the addition of new variants 
later. They are simpler, however, than object-oriented extensions, as they do not 
form subtyping between variants themselves, but only between combinations of them. 
This makes an important distinction between \emph{extensible sum types} like 
polymorphic variants and \emph{extensible hierarchical sum types} like classes.
An important property of extensible sum types is that each value of the 
underlain algebraic data type belongs to exactly one disjoint subset, tagged with 
a constructor. The \emph{nominative subtyping} of object-oriented languages does 
not usually have this disjointness making classes effectively have multiple 
types. In particular, the case of disjoint constructors can be seen as a 
degenerated case of a flat class hierarchy among the multitude of possible class 
hierarchies.

\emph{Tom} is a pattern-matching compiler that can be used together with Java, C or 
Eiffel to bring a common pattern matching and term rewriting syntax into the 
languages~\cite{Moreau:2003}. In comparison to our approach, Tom has much bigger 
goals. The combination of pattern matching, term rewriting and strategies turns 
Tom into a tree-transformation language similar to Stratego/XT, XDuce and 
others. Tom supports type patterns and its \%match statement can be used as a 
type switch; however, similarly to other object-oriented languages, Tom's 
handling of type switching is based on highly inefficient \code{instanceof} 
operator and its equivalents in other languages.

Pattern matching in Scala~\cite{Scala2nd} also allows type patterns and thus type 
switching. The language supports extensible and hierarchical data types, but 
their handling in a type switching constructs varies. Sealed classes are handled 
with an efficient switch over all tags, since sealed classes cannot be extended. 
Classes that are not sealed are similarly approached with a combination of an 
\code{InstanceOf} operator and a decision tree~\cite{EmirThesis}.
