\subsection{Solution for Tagged Classes}
\label{sec:cotc}

The memoization device outlined in \textsection\ref{sec:memdev} can, in principle, also be 
applied to tagged classes. The dynamic cast will be replaced by a small 
compile-time template meta-program that checks whether the class associated with 
the given tag is derived from the target type of the case clause. If so, a static 
cast can be used to obtain the offset.

Despite its straightforwardness, we felt that it should be possible to do better 
than the general solution, given that each class is already identified with a 
dedicated constant known at compile time.

The nominal subtyping of C++ effectively gives every class multiple types. The 
idea is thus to associate with the type not only its most derived tag, but also 
the list of tags of all its base classes. In a compiler implementation such a 
list can be stored inside the virtual table of a class, while in our library 
solution it is shared between all the instances with the same most derived tag 
in a less efficient global map, associating the tag to its tag list.

The list of tags is topologically sorted accordingly to the subtyping relation 
and terminates with a dedicated value distinct from all the tags. We call such a 
list a \emph{Tag Precedence List} (TPL) as it resembles the \emph{Class 
Precedence List} (CPL) of object-oriented descendants of Lisp (e.g. Dylan, 
Flavors, LOOPS, and CLOS) used there for \emph{linearization} of class 
hierarchies. The classes in CPL are ordered from most specific to least specific 
with siblings listed in the \emph{local precedence order} -- the order of the 
direct base classes used in the class definition. TPL is just an implementation 
detail and the only reason we distinguish TPL from CPL is that in C++ classes 
are often separated into interface and implementation classes and it might so 
happen that the same tag is associated by the user with an interface and several 
implementation classes. 

A type switch below, built on top of a hierarchy of tagged classes, proceeds as 
a regular switch on the subject's tag. If the jump succeeds, we found an exact 
match; otherwise, we get into a default clause that obtains the next tag in the 
tag precedence list and jumps back to the beginning of the switch statement for a 
rematch:

\begin{lstlisting}
    const size_t* taglist = 0;
    size_t attempt = 0; size_t tag = object->tag;
ReMatch:
    switch (tag) {
    default:
        if (!taglist) taglist = get_taglist(object->tag);
        tag = taglist[++attempt];
        goto ReMatch;
    case end_of_list: break;
    case bindings<D1>::tag_value: s1; break;
    ...
    case bindings<Dn>::tag_value: sn; break;
    }
\end{lstlisting}

\noindent
The above structure, which we call \emph{TPL Dispatcher}, lets us dispatch to case clauses of the most derived class 
with an overhead of initializing two local variables, compared to a switch on a 
sealed hierarchy. Dispatching to a case clause of a base class will take time 
roughly proportional to the distance between the matched base class and the most 
derived class in the inheritance graph. When none of the base class tags was 
matched, we will necessarily reach the end\_of\_list marker in the tag precedence 
list and thus exit the loop.

Our library automatically builds the function \code{get_taglist} based on 
additional definitions that the user provides. The details can be found in 
companion technical report~\cite[\textsection 6]{TR}.
